Introducing Distributed Dynamic Data-intensive (D3) Science: Understanding Applications and Infrastructure

### 2015-09-13 3dpas_intro.tex
removing old comments, adding a new one, fixing a bib entry

  A common feature across many science and engineering applications is the
  amount and diversity of data and computation that must be integrated to yield
  insights. Often data is increasingly large-scale and distributed. Often the
  datasets, their locations, availability, analysis and associated scheduling
  decisions are time-dependent. Collectively these give rise to dynamic
  distributed data intensive applications.  The characteristics, requirements,
  and software systems for the analysis of large volumes of ``static'' data have
  received significant attention; dynamic, distributed and data intensive have
  received relatively less. This work will ameliorate the state by surveying
  several representative application scenarios, to provide a common conceptual
  framework to understand them as well as examine the infrastructure used in
  support of applications.

@@

\ifdraft
\newpage
\tableofcontents
%\newpage
%\listoftables
\newpage
\fi

@@

\jhanote{NCH points out that we are inconsistent in the order of the
  ``D''s. SJ to fix globally. Good catch Neil!}

@@

%\ysnote{Since we're clearly positioning around BigData,
%  one question that is going to come up is how are the 3D's related to
%  the 3V's -- volume, velocity and variety -- of Big Data. We may want
%  to explicitly address that early on. For e.g. ``While the three
%  dimensions of Volume, Velocity and Variety of
%  BigData~\cite{gartner3v}, or 3V's, are well recognized in industry,
%  these narrowly consider the characteristics of the data rather than
%  the interplay between Big Data and applications that operate over
%  them as we do through the 3D's. Furthermore, our emphasis is also on
%  the role of Big Data in science, which is less studied (and arguably
%  more diverse and complex) than business applications. That said, the
%  volume and velocity dimensions of the 3V's partially capture the
%  data-intensive and dynamic aspects of our study.''}  
\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
  is assumed but not discussed}

@@

\jhanote{refine next four paragraph...}

@@

\item This {\it survey} is not meant to be complete, either in scope of domains
  covered or types of infrastructure and application surveyed.  They by
  definition represent the interests and experience of the authors.

@@

With ITER~\cite{iter} scheduled to operate before 2020, plasma fusion has become
one of the highest priority research items for the US DOE. Scalability of
first-principles fusion codes is an important part of fusion research
needed to obtain predictive capability of plasma performance with
first-principles physics fidelity. The goal is to build a realistic
full-distribution kinetic code, in realistic ITER geometry, combining
the strengths of each of the existing gyrokinetic codes (GEM, GTS,
XGC, and GTC), and utilizing the most up-to-date computer science,
hardware design, and applied mathematics technologies developed. This
{\em traditional application} is intended to be capable of 1) discovering
macroscopic magnetohydrodynamics activity in burning plasmas, 2)
understanding Energetic particle effects in fusion reactors, 3)
understanding the heating in a fusion reactor, and 4) understanding the
interaction of the plasma at the material boundary. This applications just
has the data processing stage, along with associated data movement.

@@

{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express computations over 
collections of data which are executed in a loosely-coupled, data-parallel manner.
The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
prominent among these, with recent ones like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
graph analytics. MapReduce forces programmers to model computation in two stages: 
a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
are aggregated through a binary function that is associative and commutative. 
These properties serve to guarantee that the MapReduce computation can be efficiently 
parallelized, by performing the map phase concurrently and
then reducing without the reduction order affecting the final result.

@@

In this section we attempt to summarize four key themes from previous
sections: (i) the source and nature of dynamic data that we observe in
the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
data distribution that we observe in such applications; (iii) the
general characteristics that can be associated with D3 science
applications; (iv) the characteristics of a computational
infrastructure that may be most suited for D3 science
applications. Whereas (i) and (ii) are primarily observations that we
can make on existing applications that we have discussed in
\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
trends that can be applied to other applications we have not
considered. \katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes. \katznote{not sure about the `other' part of this...}

@@

\jhanote{Note to self: need smoothening}

A survey of data-intensive distributed applications on production
cyberinfrastructure leads us to posit that there are three primary
macroscopic architectures~\cite{pilot_clouds} and approaches for distributed
data-intensive applications: (i) localize all data into a ``cloud''
(or a large analytical engine) -- the paradigm adopted by several
genome projects and the cancer atlas, (ii) decompose and distribute
data to an appropriate number of computing/analytical engines as
available -- the paradigm employed by particle physics for the
discovery of the Higgs Boson (similar to the application in
\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms,
wherein data is decomposed and committed to several infrastructure,
which in turn could be a combination of either of the first two
paradigms, similar to what is done in the Power grids application
(\S\ref{power}). \jhanote{improve description....replace hybrid by
  bottom-up? }

@@

\katznote{previous 2 paragraphs to be re-read.  Old comments about them
are available as commented out comments.}
%\jhanote{Irrespective, I find this past paragraph
%  somewhat speculative. And I don't see the point of it? WHat is the
%  latter half trying to say: there are problems with APIs and
%  interfaces of 3D applications and systems?} \katznote{It's saying
%  that applications aren't developed in a vacuum, particularly
%  infrastructural applications, and that the constraints on the
%  applications leads to new tools that are or could be developed not
%  being able to be applied in practice.  In some ways, it's a call for
%  an application abstraction or middleware, between the applications
%  and the `infrastructure' as defined in \S4, much as perhaps AIMES is
%  trying to build}\jhanote{Dan, I find your explanatory note more
%  convincing than the sentence on workflow systems! Also there is no
%  mention of middleware as the ``impedance mismatch lowering layer''
%  in the above. I propose we rewrite the sentence starting, ``Workflow
%  systems are a possible way..} \katznote{ok, I used some of my text
%  and turned around the sentence from Omer's optimistic view to
%  my more pessimistic view.  If others disagree, we can change it back
%  or compromise.}

@@

\section*{Methodology}

>>>>

  A common feature across many science and engineering applications is the
  amount and diversity of data and computation that must be integrated to yield
  insights. Often data is increasingly large-scale and distributed. Often the
  datasets, their locations, availability, analysis and associated scheduling
  decisions are time-dependent. Collectively these give rise to dynamic
  distributed data intensive applications.  The characteristics, requirements,
  and software systems for the analysis of large volumes of ``static'' data have
  received significant attention; dynamic, distributed and data intensive have
  received relatively less. This work will ameliorate the state by surveying
  several representative application scenarios, to provide a common conceptual
  framework to understand them as well as examining the infrastructure used in
  support of applications.

@@

%\ifdraft
%\newpage
%\tableofcontents
%%\newpage
%%\listoftables
%\newpage
%\fi

@@

%\jhanote{NCH points out that we are inconsistent in the order of the
%  ``D''s. SJ to fix globally. Good catch Neil!}

@@

%\ysnote{Since we're clearly positioning around BigData,
%  one question that is going to come up is how are the 3D's related to
%  the 3V's -- volume, velocity and variety -- of Big Data. We may want
%  to explicitly address that early on. For e.g. ``While the three
%  dimensions of Volume, Velocity and Variety of
%  BigData~\cite{gartner3v}, or 3V's, are well recognized in industry,
%  these narrowly consider the characteristics of the data rather than
%  the interplay between Big Data and applications that operate over
%  them as we do through the 3D's. Furthermore, our emphasis is also on
%  the role of Big Data in science, which is less studied (and arguably
%  more diverse and complex) than business applications. That said, the
%  volume and velocity dimensions of the 3V's partially capture the
%  data-intensive and dynamic aspects of our study.''}  
%\jhanote{we do not discuss explicitly the 4th D, i.e. diversity (courtesy AL) it
%  is assumed but not discussed}

@@

%\jhanote{refine next four paragraph...}

@@

\item This {\it survey} is not meant to be complete, either in scope of domains
  covered or types of infrastructure and application surveyed.  They by
  definition represent the interests and experience of the authors.
  \katznote{need to add a note about the dates in which application data was collected?  I'm not sure the numbers are correct today, for example, from ATLAS}

@@

With ITER~\cite{iter} scheduled to operate perhaps around 2022, plasma fusion has become
one of the highest priority research items for the US DOE. Scalability of
first-principles fusion codes is an important part of fusion research
needed to obtain predictive capability of plasma performance with
first-principles physics fidelity. The goal is to build a realistic
full-distribution kinetic code, in realistic ITER geometry, combining
the strengths of each of the existing gyrokinetic codes (GEM, GTS,
XGC, and GTC), and utilizing the most up-to-date computer science,
hardware design, and applied mathematics technologies developed. This
{\em traditional application} is intended to be capable of 1) discovering
macroscopic magnetohydrodynamics activity in burning plasmas, 2)
understanding Energetic particle effects in fusion reactors, 3)
understanding the heating in a fusion reactor, and 4) understanding the
interaction of the plasma at the material boundary. This applications just
has the data processing stage, along with associated data movement.

@@

{\em Bulk Data-parallel} frameworks use ``Bulk'' operators to express computations over 
collections of data which are executed in a loosely-coupled, data-parallel manner.
The {\em MapReduce} paradigm~\cite{MapReduce} and its variants~\cite{twister,hmr} are the most
prominent among these, with those like Pregel and Apace Giraph~\cite{pregel} supporting specific domains like
graph analytics. MapReduce forces programmers to model computation in two stages: 
a \emph{map} phase in which a transformation is applied to every element of a collection independently of every other;  and a \emph{reduce} phase in which the transformed results 
are aggregated through a binary function that is associative and commutative. 
These properties serve to guarantee that the MapReduce computation can be efficiently 
parallelized, by performing the map phase concurrently and
then reducing without the reduction order affecting the final result.

@@

In this section we attempt to summarize four key themes from previous
sections: (i) the source and nature of dynamic data that we observe in
the application surveyed in \S\ref{sec:scenarios}; (ii) the types of
data distribution that we observe in such applications; (iii) the
general characteristics that can be associated with D3 science
applications; (iv) the characteristics of a computational
infrastructure that may be most suited for D3 science
applications. Whereas (i) and (ii) are primarily observations that we
can make on existing applications that we have discussed in
\S\ref{sec:scenarios}, (iii) and (iv) attempt to synthesize general
trends that can be applied to other applications we have not
considered. 
%\katznote{should this be split into two subsections then? Perhaps Observations \& Trends?}

@@

From \S\ref{sec:distTypes} and
\figurename~\ref{fig:figures_application}, we find that applications
with distributed data differ based on the amount or volume of data
distributed, the degree of distribution (scattered over one site or
many sites), and the dynamic aspects associated with distribution,
e.g., latency and relative scheduling flexibility (and thus tolerance
or constraints on data transfer/aggregation).  Applications processing
the largest volumes of data (O(PB)) typically run on specialized data
cyberinfrastructure (e.g., the Atlas LHC Grid). Applications such as
Fusion and NGS that run on smaller data sets mainly utilize HPC
environments. On the other end of the spectrum are highly dynamic
applications that process sensor data and make near-realtime
decisions based on streaming data, e.g., the power grid and sensor
network applications. These applications process much less data;
often, sensor data is also archived and aggregated for historical
analyses, which leads to higher data volumes. %\katznote{not sure about the `other' part of this...}

@@

%\jhanote{Note to self: need smoothening}

A survey of data-intensive distributed applications on production
cyberinfrastructure leads us to posit that there are three primary
macroscopic architectures~\cite{pilot_clouds} and approaches for distributed
data-intensive applications: (i) localize all data into a ``cloud''
(or a large analytical engine) -- the paradigm adopted by several
genome projects and the cancer atlas, (ii) decompose and distribute
data to an appropriate number of computing/analytical engines as
available -- the paradigm employed by particle physics for the
discovery of the Higgs Boson (similar to the application in
\S\ref{WLCGSteve}), and (iii) a hybrid of the above two paradigms,
wherein data is decomposed and committed to several infrastructure,
which in turn could be a combination of either of the first two
paradigms, similar to what is done in the Power grids application
(\S\ref{power}). %\jhanote{improve description....replace hybrid by bottom-up? }

@@

%\katznote{previous 2 paragraphs to be re-read.  Old comments about them
%are available as commented out comments.}
%\jhanote{Irrespective, I find this past paragraph
%  somewhat speculative. And I don't see the point of it? WHat is the
%  latter half trying to say: there are problems with APIs and
%  interfaces of 3D applications and systems?} \katznote{It's saying
%  that applications aren't developed in a vacuum, particularly
%  infrastructural applications, and that the constraints on the
%  applications leads to new tools that are or could be developed not
%  being able to be applied in practice.  In some ways, it's a call for
%  an application abstraction or middleware, between the applications
%  and the `infrastructure' as defined in \S4, much as perhaps AIMES is
%  trying to build}\jhanote{Dan, I find your explanatory note more
%  convincing than the sentence on workflow systems! Also there is no
%  mention of middleware as the ``impedance mismatch lowering layer''
%  in the above. I propose we rewrite the sentence starting, ``Workflow
%  systems are a possible way..} \katznote{ok, I used some of my text
%  and turned around the sentence from Omer's optimistic view to
%  my more pessimistic view.  If others disagree, we can change it back
%  or compromise.}

@@

\appendix
\section*{Appendix: Methodology}
